HTT_Leavitt HTT mRNA Expression in Cortex vs Liver
================================================================
##### By: Sumaiya Islam
##### Date: Sept 1, 2015

### Script contents:
- Test for significance of difference of averaged HTT mRNA expression measures (from 3 replicates in each tissue) between cortex vs liver of 4 individuals from matched 450K dataset

### A. Set up wd & packages

We will initially set our working directory and load our libraries.
```{r HTT_Leavitt_setup, include = FALSE, echo=FALSE, warning=FALSE}
setwd("/home/sislam/HTT_Leavitt")
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library("RColorBrewer")
library(knitr)
library(xtable)
library(limma)
library(reshape)
library(ggplot2)
library(grid)
library(gridExtra)
library(MASS)
```

### B. Load data

We will load the data containing the averaged HTT mRNA expression values (from 3 replicates from each tissue) in cortex and liver. 

```{r load_data, echo=FALSE, warning=FALSE}
HTT.mRNA.dat<-read.csv(file="HTT_Tissues_Expression.csv", header = T, check.names=F)
head(HTT.mRNA.dat)
str(HTT.mRNA.dat)
```

### C. Check for normality

Note that interpretation of normality test is adapted from: http://stats.stackexchange.com/questions/15696/interpretation-of-shapiro-test


```{r normality_check, echo=FALSE}
(Liver.normality<-shapiro.test(HTT.mRNA.dat$Liver))
(Cortex.normality<-shapiro.test(HTT.mRNA.dat$Cortex))
```
We can see that since the Shapiro-Wilk test for normality yielded a p-value > 0.05 for both the Cortex and Liver measures. We can conclude that we cannot reject the hypothesis that the sample measures come from a population which has a normal distribution.

Also according to the q-plot for the Cortex and Liver measures:


```{r qplot, fig.height=3, fig.width=3, echo=FALSE}
qqnorm(HTT.mRNA.dat$Liver)
qqnorm(HTT.mRNA.dat$Cortex)
```


The qqplots for Cortex and Liver data are relatively normal.

##### Since the Shapiro test doesn't indicate that the data come from a population with a non-normal distribution and the data looks vaguely normal according to q-q plots, we can assume that the measures likely arise from a Gaussian distribution. Given that t-tests are quite robust to violations of normality assumption, it seems that a paired t-test would be appropriate in this case.

### D. Test for significance of difference between Cortex and Liver HTT expression values

```{r stats_test, fig.height=3, fig.width=3, echo=FALSE}

head(melted.dat<-melt(HTT.mRNA.dat))
ggplot(melted.dat, aes(variable, value))+
  geom_boxplot(color="grey", outlier.shape=NA)+
  geom_point(color = "#3182bd", shape=19, size=4, position = position_jitter(w = 0.05))+
  theme_bw()+xlab("Tissue")+ylab("Averaged HTT mRNA Expression")

(stat.test<-t.test(HTT.mRNA.dat$Cortex, HTT.mRNA.dat$Liver, paired=T))
```

We can say that the difference in the averaged HTT mRNA expression measures are statistically significant (paired t-test, p = 0.023)

