HTT_Leavitt PCA & ComBat
========================================================
## By: Sumaiya Islam
## Date: August 1, 2015 (note: script was reviewed and updated on March 7, 2016)

### Script contents:
- PCA comparing the raw and normalized datasets for cortex-only dataset 
- ComBat to correct technical batch effects for cortex-only dataset 
##### Note: Since the HTT_Leavitt matched dataset was run on a single chip, in the same batch, we do not need to perform ComBat to correct for technical variation in the matched dataset. However, since the cortex-only dataset was run across two chips, but in the same batch, we do need to perform ComBat to correct for technical variation. 

### A. Set up working directory & packages


We will initially set our working directory and load our libraries.
```{r}
library(methylumi)
library(reshape)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(gridExtra)
library(sva)
library(rama)
```

## Heat scree plot Function
```{r}
### Function of association meta variable with PC (ANOVA)
heat_scree_plot<-function(Loadings, Importance, Num, Order){
  adjust<-1-Importance[1]
  pca_adjusted<-Importance[2:length(Importance)]/adjust
  pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted)))
  
  scree<-ggplot(pca_df[which(pca_df$PC<Num),],aes(PC,adjusted_variance))+geom_bar(stat = "identity",color="black",fill="grey")+theme_bw()+
        theme(axis.text = element_text(size =12),
              axis.title = element_text(size =15),
              plot.margin=unit(c(1,1.5,0.2,2.25),"cm"))+ylab("Variance")+
    scale_x_continuous(breaks = seq(1,Num,1))
  
  #### Heat
  ## correlate meta with PCS
  ## Run anova of each PC on each meta data variable

  aov_PC_meta<-lapply(1:ncol(meta_categorical), function(covar) sapply(1:ncol(Loadings), function(PC) summary(aov(Loadings[,PC]~meta_categorical[,covar]))[[1]]$"Pr(>F)"[1]))
   cor_PC_meta<-lapply(1:ncol(meta_continuous), function(covar) sapply(1:ncol(Loadings), function(PC) (cor.test(Loadings[,PC],as.numeric(meta_continuous[,covar]),alternative = "two.sided", method="spearman", na.action=na.omit)$p.value)))
  names(aov_PC_meta)<-colnames(meta_categorical)
  names(cor_PC_meta)<-colnames(meta_continuous)
  aov_PC_meta<-do.call(rbind, aov_PC_meta)
  cor_PC_meta<-do.call(rbind, cor_PC_meta)
  aov_PC_meta<-rbind(aov_PC_meta, cor_PC_meta)
  aov_PC_meta<-as.data.frame(aov_PC_meta)
  #adjust
  aov_PC_meta_adjust<-aov_PC_meta[,2:ncol(aov_PC_meta)]
  
    
  #reshape
  avo<-aov_PC_meta_adjust[,1:(Num-1)]
  avo_heat_num<-apply(avo,2, as.numeric)
  avo_heat<-as.data.frame(avo_heat_num)
  colnames(avo_heat)<-sapply(1:(Num-1), function(x) paste("PC",x, sep=""))
  avo_heat$meta<-rownames(avo)
  avo_heat_melt<-melt(avo_heat, id=c("meta"))
  
  # cluster meta data
  ord <- Order
  meta_var_order<-unique(avo_heat_melt$meta)[rev(ord)]
  avo_heat_melt$meta <- factor(avo_heat_melt$meta, levels = meta_var_order)
  
  # color if sig
  # avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]>=0.9){">=0.9"}else{
   # if(avo_heat_melt$value[x]>=0.5){">=0.5"}else{
     # if(avo_heat_melt$value[x]>=0.1){">=0.1"}else{"<0.1"}}})
  avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]<=0.001){"<=0.001"}else{
     if(avo_heat_melt$value[x]<=0.01){"<=0.01"}else{
       if(avo_heat_melt$value[x]<=0.05){"<=0.05"}else{">0.05"}}})
  
  heat<-ggplot(avo_heat_melt, aes(variable,meta, fill = Pvalue)) +
  geom_tile(color = "black",size=0.5) +
  theme_gray(8)+scale_fill_manual(values=c("#084594","#4292c6","#9ecae1","#deebf7"))+
      theme(axis.text = element_text(size =10, color="black"),
            axis.text.x = element_text(),
          axis.title = element_text(size =15),
          legend.text = element_text(size =14),
          legend.title = element_text(size =12),
          legend.position = c(1, 0), legend.justification = c(1,0),
          plot.margin=unit(c(0,2.25,1,1),"cm"))+
    xlab("Principal Component")+ylab(NULL)
  
  grid.arrange(scree, heat, ncol=1)
}
```


### B. Load files

#### We will be analyzing the raw and normalized datasets for the cortex-only dataset

```{r}

load("HTT_Leavitt.cortex.norm.RData") # normalized data for cortex-only dataset
# check for NAs in data
ind<-is.row.na(betas(HTT_Leavitt.cortex.swan)) # The function returns a vector of logical variables, one for each row of the matrix. The variable is TRUE if the row does not contain any missing values and FAlSE otherwise.
length(na.count<-which(ind=="FALSE")) # 0 (there are no rows that contain a NA in the raw data)
```


## PCA Scree Heatmap for raw data for cancer.dat

```{r warning=FALSE, fig.height=9, fig.width=11}

## re-structure meta data: change categorical variables to factors for ANOVA and continuous variables to numeric for Spearman's correlation
meta.cortex<-pData(HTT_Leavitt.cortex.swan)
str(meta.cortex)

## PCA 
betas.cortex<-betas(HTT_Leavitt.cortex.swan)
PCA_full<-princomp(betas.cortex[complete.cases(betas.cortex),]) # scaling is not necessary for normalized dataset
Loadings<-as.data.frame(unclass(PCA_full$loadings))                                                                                                                                                                                                                vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)
adjust<-1-Importance[1]
pca_adjusted<-Importance[2:length(Importance)]/adjust
(pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted))))


#Specify which covariates are categorical and/or categorical
# structure meta data appropriately
colnames(meta.cortex)
meta.cortex$Sentrix_ID <- as.factor(meta.cortex$Sentrix_ID)
meta.cortex$Sentrix_Position <- as.character(meta.cortex$Sentrix_Position)
meta.cortex$Sample_Group <- as.factor(meta.cortex$Sample_Group)
meta.cortex$Sex<- as.factor(meta.cortex$Sex)
meta.cortex$Disease_status <- as.factor(meta.cortex$Disease_status)
meta.cortex$HD_CAG_length <- as.numeric(meta.cortex$HD_CAG_length) 
meta.cortex$WT_CAG_length <- as.numeric(meta.cortex$WT_CAG_length)
meta.cortex$Age_Death <- as.numeric(meta.cortex$Age_Death)
meta.cortex$Age_Onset  <- as.numeric(meta.cortex$Age_Onset)
# for Row
for (i in 1:nrow(meta.cortex)){
  meta.cortex$Row[i]<-paste(substr(meta.cortex[i,"Sentrix_Position"], start=1, stop=3))
}
meta.cortex$Row <- as.factor(meta.cortex$Row)
str(meta.cortex)

# pull out variables to associate with PCs
meta_categorical<-meta.cortex[,c("Disease_status", "Sex", "Sentrix_ID", "Row")]  # input column numbers in meta that contain categorical variables
meta_continuous<-meta.cortex[,c("HD_CAG_length", "WT_CAG_length",  "Age_Death", "Age_Onset")] # input column numbers in meta that contain continuous variables
meta_continuous<-data.frame(meta_continuous)
colnames(meta_categorical)<-c("Disease status","Sex", "Chip", "Row")
colnames(meta_continuous)<-c("HD CAG length", "WT CAG length", "Age of Death", "Age of Disease Onset")

# Specifiy the number of PCs you want shown
Num<-13

# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-1:8

#Apply function on PCA results, pulls in the meta data and beta values from above
heat_scree_plot(Loadings, Importance, Num, Order)                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                     
```


We can observe that there are some chip-to-chip effects since chip is correlated with PC9. Accordingly, we need to correct for chip using Combat. There does not appear to be any row effects so we will not correct for row using Combat.

Note that there is there is ~30% of the DNA methylation variation in PC1 that is not accounted for by any of the variables listed here. This 30% variance is likely due to brain cell type proportion differences which we will investigate using a brain cell type deconvolution method. Also note that age of death correlates with PC2 and sex is correlated with PC4. We will look more closely at the association of these biological variables to PCs after cell type correction. 

### B. ComBat 

We will apply ComBat to the normalized dataset containing the cortex samples only (which were run across the two batches). ComBat uses an empirical Bayesian (EB) estimates the Location(mean)/Scale(variance) model parameters that represent the batch effects by “pooling information” across genes in each batch to “shrink” the batch effect parameter estimates toward the overall mean of the batch effect estimates (across genes). These EB estimates are then used to adjust the data for batch effects, providing more robust adjustments for the batch effect on each gene.

According to the creators of the SVA package, we can pre-set which variable(s) of interest to protect from ComBat correction. We will only protect Disease_status as that is our main variable of interest. We will specifically correct for chip effects.


```{r echo=FALSE}

preComBat_Batch.datM <- exprs(HTT_Leavitt.cortex.swan) # ComBat must be applied to M-values from normalized dataset
mod <- model.matrix(~ Disease_status, data=meta.cortex)
Combat_Batch.dat <- ComBat(dat=preComBat_Batch.datM, batch=meta.cortex$Sentrix_ID, mod=mod)


### fill in the slot of the methylumi object with ComBat file
Combat_1<-HTT_Leavitt.cortex.swan
identical(rownames(exprs(Combat_1)), rownames(Combat_Batch.dat))
exprs(Combat_1) <- Combat_Batch.dat

# pull needed data for PCA from ComBat 1 methylumi object
Beta_Data<-betas(Combat_1)
```

### C. PCA for Combat (Batch effect corrected) Data

## Run PCA and apply the heat_scree plot function
```{r warning=FALSE, fig.height=9, fig.width=11}
## PCA
PCA_full<-princomp(Beta_Data[complete.cases(Beta_Data),]) # scaling is not necessary for normalized dataset
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)
adjust<-1-Importance[1]
pca_adjusted<-Importance[2:length(Importance)]/adjust
(pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted))))

#Specify which covariates are categorical and/or categorical
meta_categorical<-meta.cortex[,c("Disease_status", "Sex", "Sentrix_ID", "Row")]  # input column numbers in meta that contain categorical variables
meta_continuous<-meta.cortex[,c("HD_CAG_length", "WT_CAG_length",  "Age_Death", "Age_Onset")] # input column numbers in meta that contain continuous variables
meta_continuous<-data.frame(meta_continuous)
colnames(meta_categorical)<-c("Disease status","Sex", "Chip", "Row")
colnames(meta_continuous)<-c("HD CAG length", "WT CAG length", "Age of Death", "Age of Disease Onset")

# Specifiy the number of PCs you want shown
Num<-13

# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-1:8

#Apply function on PCA results, pulls in the meta data and beta values from above
heat_scree_plot(Loadings, Importance, Num, Order) 
```

We can see that chip effects have been largely removed from the methylation data. (ie only associate with last PC accounting for only 1.98% of the variance). No row effects are observed.  

### D. Save Combat corrected data
```{r}
HTT_Leavitt.cortex.combat_copy<-Combat_1
dim(HTT_Leavitt.cortex.combat_copy) # probes = 437,035 n = 13
# save(HTT_Leavitt.cortex.combat_copy, file="HTT_Leavitt.cortex.combat_copy.RData")
colnames(exprs(HTT_Leavitt.cortex.combat_copy))
```

We will carry on with cell type correction of the cortex samples in the next script. 