HTT_Leavitt Pyrosequencing of CTCF Binding Site in HTT Promoter
================================================================
## By: Sumaiya Islam
## Date: April 21, 2015

### Script contents:
- Generate heatmap of the methylation values (measured by pyrosequencing) for CpGs at CTCF binding site in HTT promoter

### A. Set up wd & packages

We will initially set our working directory and load our libraries.
```{r HTT_Leavitt_setup, include = FALSE, echo=FALSE}
setwd("/home/sislam/HTT_Leavitt")
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library("RColorBrewer")
library(knitr)
library(xtable)
library(limma)
library(reshape)
library(ggplot2)
library(grid)
library(gridExtra)
library(MASS)
```

### B. Load data

We will load the data containing both the methylation values for the 3 sites (position 1, 2, 3) in the CTCF binding site

```{r load_data}
CTCF.pyro.dat<-read.csv(file="CTCF_Binding_Site_Pyro_Results.csv", header = T, check.names=F)
head(CTCF.pyro.dat)
str(CTCF.pyro.dat)
CTCF.pyro.dat$Sample_Name<-as.character(CTCF.pyro.dat$Sample_Name)
CTCF.pyro<-CTCF.pyro.dat[which(CTCF.pyro.dat$Tissue%in%c("Liver", "Cortex")),]
str(CTCF.pyro)


# subset out matrix of data points and scale data
CTCF.pyro.data<-CTCF.pyro[,7:9]
rownames(CTCF.pyro.data)<-CTCF.pyro$Sample_Name
CTCF.pyro.dat.scaled<-scale(CTCF.pyro.data)
```

### C. Heatmaps

#### Heatmap of all samples pyrosequenced
```{r}
# # colour key for tissue 
# tissueCol <- as.numeric(factor(CTCF.pyro$Tissue))
# tissueCol <- gsub("1", "orange",  gsub("2", "blue", gsub("3", "forestgreen", tissueCol)))
# 
# # colour palette
# BuPu <- colorRampPalette(brewer.pal(n = 9, "BuPu"))
# 
# # heatmap
# heatmap.2(CTCF.pyro.dat.scaled,dendrogram="row", symm=FALSE, Rowv=TRUE, Colv=FALSE, trace = "none", col = BuPu(256), RowSideColors = tissueCol, cexCol = 0.8, margins=c(9,9))
# legend("top", legend = c("Cortex", "Liver", "Testes"),fill = c("orange","blue", "forestgreen"), horiz = T)
```

#### Heatmap of samples of original dataset 
```{r}
# # subset out samples from original dataset
# original.samples.dat<-subset(CTCF.pyro, Dataset=="Original", select=c("Sample_Name","Tissue", "Position_1", "Position_2", "Position_3"))
# str(original.samples.dat)
# orig.dat<-original.samples.dat[,3:5]
# rownames(orig.dat)<-original.samples.dat$Sample_Name
# orig.dat.scaled<-scale(orig.dat)
# 
# # colour key for tissue 
# tissueCol <- as.numeric(factor(original.samples.dat$Tissue))
# tissueCol <- gsub("1", "orange",  gsub("2", "blue", gsub("3", "forestgreen", tissueCol)))
# 
# # colour palette
# BuPu <- colorRampPalette(brewer.pal(n = 9, "BuPu"))
# 
# # heatmap
# heatmap.2(orig.dat.scaled,dendrogram="row", symm=FALSE, Rowv=TRUE, Colv=FALSE, trace = "none", col = BuPu(256), RowSideColors = tissueCol, cexCol = 0.8, margins=c(9,9))
# legend("top", legend = c("Cortex", "Liver", "Testes"),fill = c("orange","blue", "forestgreen"), horiz = T)
```

#### Heatmap of samples of validation dataset 
```{r}
# # subset out samples from original dataset
# validation.samples.dat<-subset(CTCF.pyro, Dataset=="Validation", select=c("Sample_Name","Tissue", "Position_1", "Position_2", "Position_3"))
# str(validation.samples.dat)
# vali.dat<-validation.samples.dat[,3:5]
# rownames(vali.dat)<-validation.samples.dat$Sample_Name
# vali.dat.scaled<-scale(vali.dat)
# head(vali.dat.scaled)
# 
# # colour key for tissue 
# tissueCol <- as.numeric(factor(validation.samples.dat$Tissue))
# tissueCol <- gsub("1", "orange",  gsub("2", "blue", gsub("3", "forestgreen", tissueCol)))
# 
# # colour palette
# BuPu <- colorRampPalette(brewer.pal(n = 9, "BuPu"))
# 
# # heatmap
# heatmap.2(vali.dat.scaled,dendrogram="row", symm=FALSE, Rowv=TRUE, Colv=FALSE, trace = "none", col = BuPu(256), RowSideColors = tissueCol, cexCol = 0.8, margins=c(9,9))
# legend("top", legend = c("Cortex", "Liver", "Testes"),fill = c("orange","blue", "forestgreen"), horiz = T)
```

#### Heatmap of liver and cortex samples of original dataset 
```{r}
# subset out samples from original dataset
original.corliv.dat<-CTCF.pyro[which(CTCF.pyro$Dataset=="Original" & CTCF.pyro$Tissue%in%c("Liver", "Cortex")),]
colnames(original.corliv.dat)
orig.corliv.dat<-original.corliv.dat[,7:9]
rownames(orig.corliv.dat)<-original.corliv.dat$Sample_Name
orig.corliv.dat.scaled<-scale(orig.corliv.dat)


# colour key for tissue 
tissueCol <- as.numeric(factor(original.corliv.dat$Tissue))
tissueCol <- gsub("1", "orange",  gsub("2", "blue", tissueCol))

# colour palette
BuPu <- colorRampPalette(brewer.pal(n = 9, "BuPu"))

# heatmap
heatmap.2(orig.corliv.dat.scaled,dendrogram="row", symm=FALSE, Rowv=TRUE, Colv=FALSE, trace = "none", col = BuPu(256), RowSideColors = tissueCol, cexCol = 0.8, margins=c(9,9))
legend("top", legend = c("Cortex", "Liver"),fill = c("orange","blue"), horiz = T)
```

#### Heatmap of liver and cortex samples of validation dataset 
```{r}
# subset out samples from original dataset
validation.corliv.dat<-CTCF.pyro[which(CTCF.pyro$Dataset=="Validation" & CTCF.pyro$Tissue%in%c("Liver", "Cortex")),]
colnames(validation.corliv.dat)
vali.corliv.dat<-validation.corliv.dat[,7:9]
rownames(vali.corliv.dat)<-validation.corliv.dat$Sample_Name
vali.corliv.dat.scaled<-scale(vali.corliv.dat)


# colour key for tissue 
tissueCol <- as.numeric(factor(validation.corliv.dat$Tissue))
tissueCol <- gsub("1", "orange",  gsub("2", "blue", tissueCol))

# colour palette
BuPu <- colorRampPalette(brewer.pal(n = 9, "BuPu"))

# heatmap
heatmap.2(vali.corliv.dat.scaled,dendrogram="row", symm=FALSE, Rowv=TRUE, Colv=FALSE, trace = "none", col = BuPu(256), RowSideColors = tissueCol, cexCol = 0.8, margins=c(9,9))
legend("top", legend = c("Cortex", "Liver"),fill = c("orange","blue"), horiz = T)
```

#### Generate Boxplots for CTCF Pyrosequencing Measures and Wilcoxon signed-rank (paired) testing
```{r}
# check for normality
CpG1.normality<-shapiro.test(CTCF.pyro$Position_1)
CpG1.normality$p.value # p = 0.012, p-value is lower than 0.05, can conclude that data deviates from normality
(CpG1.test<-wilcox.test(CTCF.pyro$Position_1[1:7], CTCF.pyro$Position_1[8:14], paired=TRUE)) # p = 0.016
ggplot(CTCF.pyro, aes(Tissue, Position_1))+
  geom_boxplot(color="grey", outlier.shape=NA)+
  geom_point(aes(color=Dataset), shape=19, size=4, position = position_jitter(w = 0.2))+
  theme_bw()+xlab("Tissue")+ylab("CpG 1 Methylation (%)")+
  scale_color_manual(values=c("#99d594", "#fc8d59"))

CpG2.normality<-shapiro.test(CTCF.pyro$Position_2)
CpG2.normality$p.value # p = 0.005, p-value is lower than 0.05, can conclude that data deviates from normality
(CpG2.test<-wilcox.test(CTCF.pyro$Position_2[1:7], CTCF.pyro$Position_2[8:14], paired=TRUE)) # p = 0.016
ggplot(CTCF.pyro, aes(Tissue, Position_2))+
  geom_boxplot(color="grey", outlier.shape=NA)+
  geom_point(aes(color=Dataset), shape=19, size=4, position = position_jitter(w = 0.2))+
  theme_bw()+xlab("Tissue")+ylab("CpG 2 Methylation (%)")+
  scale_color_manual(values=c("#99d594", "#fc8d59"))

CpG3.normality<-shapiro.test(CTCF.pyro$Position_3)
CpG3.normality$p.value # p = 0.026, p-value is lower than 0.05, can conclude that data deviates from normality
(CpG3.test<-wilcox.test(CTCF.pyro$Position_3[1:7], CTCF.pyro$Position_3[8:14], paired=TRUE)) # p = 0.016
ggplot(CTCF.pyro, aes(Tissue, Position_3))+
  geom_boxplot(color="grey", outlier.shape=NA)+
  geom_point(aes(color=Dataset), shape=19, size=4, position = position_jitter(w = 0.2))+
  theme_bw()+xlab("Tissue")+ylab("CpG 3 Methylation (%)")+
  scale_color_manual(values=c("#99d594", "#fc8d59"))


```

