HTT_Leavitt Epigenetic Age Analysis
====================================================================
## By: Sumaiya Islam
## Date: June 15, 2015

### Script contents:
- Assessment of 'epigenetic age' (Horvath et al. 2013, Genome Biol) to see if DNA methylation age differs between cortex vs. liver of HD cases vs. controls

### A. Set up wd & packages

We will initially set up our working directory and load our libraries.
```{r include=FALSE}
setwd("/home/sislam/HTT_Leavitt")
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library("RColorBrewer")
library(knitr)
library(xtable)
library(limma)
library(ggplot2)
library(reshape)
```

### B. Load data

We will load the raw data (un-normalized) for the cortex-only and matched datasets

```{r load_data}

## Generate raw complete methylumi object from Genome Studio exports
# get GS exports
sampleFile <- ("HTT_Leavitt_samplesheet_corrected.txt")  
allFile <- ("HTT_Leavitt.alldata.GS.txt") 
betaFile <- ("HTT_Leavitt.avgbeta.GS.txt") 
qcFile <- ("HTT_Leavitt.qc.GS.txt") 
file.exists(allFile)
file.exists(betaFile)
file.exists(qcFile)
file.exists(sampleFile)
# generate methylumi object
HTT_Leavitt<- lumiMethyR(allFile) # probes = 485,577, n = 18
# input meta data
str(sampleInfo <- read.delim(sampleFile,row.names=1,stringsAsFactors=F))
stopifnot(all(sampleNames(HTT_Leavitt)%in%rownames(sampleInfo)))
sampleInfo <- sampleInfo[sampleNames(HTT_Leavitt),]
# re-structure meta data
str(pData(HTT_Leavitt) <- sampleInfo)
pData(HTT_Leavitt)$Sample_Group<- as.factor(pData(HTT_Leavitt)$Sample_Group)
pData(HTT_Leavitt)$Sex<- as.factor(pData(HTT_Leavitt)$Sex)
pData(HTT_Leavitt)$Disease_status<- as.factor(pData(HTT_Leavitt)$Disease_status)
pData(HTT_Leavitt)$Age_Death<- as.numeric(pData(HTT_Leavitt)$Age_Death)
str(pData(HTT_Leavitt))
identical(rownames(pData(HTT_Leavitt)), colnames(exprs(HTT_Leavitt))) # TRUE
sampleNames(HTT_Leavitt)<-c("COB05_Cortex", "HDB176_Cortex","COB_22_52_Cortex", "HDB167_Cortex","HDB162_Cortex", "COB51_Cortex",     "HDB119_Cortex", "HDB166_Cortex",  "COB125_Cortex","COB20_30_Cortex","COB59_Cortex",   "HDB159_Cortex","HDB165_Cortex","COB51_Liver",    
"HDB119_Liver",  "HDB162_Liver",  "HDB176_Liver", "HDB167_Liver") 
identical(sampleNames(HTT_Leavitt), rownames(pData(HTT_Leavitt))) # TRUE

## generate cortex-only dataset
# subset cortex samples
cortex.samples<- grep("Cortex", HTT_Leavitt$Sample_Group)
HTT_Leavitt.cortex<-HTT_Leavitt[,cortex.samples]
dim(HTT_Leavitt.cortex) # probes = 485,577  n = 13
meta.cortex<-pData(HTT_Leavitt.cortex)
head(betas.cortex<-betas(HTT_Leavitt.cortex))
# write.csv(betas.cortex, file="Betas_cortex.csv")



## generate matched dataset
matched.samples<- c("HDB119_Cortex", "HDB119_Liver", "COB51_Cortex", "COB51_Liver", "HDB167_Cortex", "HDB167_Liver", "HDB176_Cortex", "HDB176_Liver", "HDB162_Cortex", "HDB162_Liver")
HTT_Leavitt.Matched<-HTT_Leavitt[,matched.samples] # probes= 437035, n = 10
dim(HTT_Leavitt.Matched) # probes = 485,577  n = 10
meta.matched<-pData(HTT_Leavitt.Matched)
head(betas.matched<-betas(HTT_Leavitt.Matched))
# write.csv(betas.matched, file="Betas_matched.csv")
```

### C. Re-shape beta data put input into epigenetic age calculator

```{r}
# re-format cortex-only dataset
library(sqldf)
#change the setwd filepath to that of the folder with your data. Note the forward slash
setwd("/home/sislam/HTT_Leavitt")
betas.cortex.df<-as.data.frame(betas.cortex)
Probe_ID<-rownames(betas.cortex) # character vector of length 485,577
dat0<-cbind(Probe_ID,betas.cortex.df)
str(dat0)
datMiniAnnotation=read.csv("datMiniAnnotation.csv")
match1=match(datMiniAnnotation[,1], dat0[,1])
dat0Reduced=dat0[match1,]
head(dat0Reduced)
dat0Reduced[,1]=as.character(dat0Reduced[,1])
dat0Reduced[is.na(match1),1]=as.character(datMiniAnnotation[is.na(match1),1])
datout=data.frame(dat0Reduced)
# make sure you output numeric variables...
for (i in 2:dim(datout)[[2]]  ){datout[,i]= 
as.numeric(as.character(gsub(x=datout[,i],pattern="\"",replacement=""))) }
head(datout)
str(datout)
# write.table(datout,"Betas.cortex.Horvath.csv", row.names=F, sep="," )

# re-format matched dataset
betas.matched.df<-as.data.frame(betas.matched)
Probe_ID<-rownames(betas.matched) # character vector of length 485,577
dat0<-cbind(Probe_ID,betas.matched.df)
str(dat0)
datMiniAnnotation=read.csv("datMiniAnnotation.csv")
match1=match(datMiniAnnotation[,1], dat0[,1])
dat0Reduced=dat0[match1,]
head(dat0Reduced)
dat0Reduced[,1]=as.character(dat0Reduced[,1])
dat0Reduced[is.na(match1),1]=as.character(datMiniAnnotation[is.na(match1),1])
datout=data.frame(dat0Reduced)
# make sure you output numeric variables...
for (i in 2:dim(datout)[[2]]  ){datout[,i]= 
as.numeric(as.character(gsub(x=datout[,i],pattern="\"",replacement=""))) }
head(datout)
str(datout)
#replace "MethylationData" with a filename of your choice
# write.table(datout,"Betas.matched.Horvath.csv", row.names=F, sep="," )
```

### D. Select needed variables for meta data to input into epigenetic age calculator
```{r}
## for cortex only dataset meta data
colnames(meta.cortex)
meta.cortex$Sample_Name<-rownames(meta.cortex)
meta.cortex.Horvath<-subset(meta.cortex, select=c("Sample_Name","Age_Death", "Sex", "Sample_Group"))
# re-format data according to Horvath's specifications in http://labs.genetics.ucla.edu/horvath/dnamage/TUTORIALonlineCalculator.pdf
colnames(meta.cortex.Horvath)<-c("Sample_ID","Age", "Female", "Tissue")
meta.cortex.Horvath$Female<-gsub("F", "1", gsub("M", "0", meta.cortex.Horvath$Female))
meta.cortex.Horvath$Female<-as.numeric(meta.cortex.Horvath$Female)
meta.cortex.Horvath$Tissue<-"Brain FCTX" 
identical(colnames(betas.cortex), meta.cortex.Horvath$Sample_ID) #TRUE
meta.cortex.Horvath
# write.table(meta.cortex.Horvath,"Meta.cortex.Horvath.csv", row.names=F, sep=",")

## for matched dataset meta data
colnames(meta.matched)
meta.matched.Horvath<-subset(meta.matched, select=c("Sample_Name","Age_Death", "Sex", "Sample_Group"))
# re-format data according to Horvath's specifications in http://labs.genetics.ucla.edu/horvath/dnamage/TUTORIALonlineCalculator.pdf
colnames(meta.matched.Horvath)<-c("Sample_ID","Age", "Female", "Tissue")
meta.matched.Horvath$Female<-gsub("F", "1", gsub("M", "0", meta.matched.Horvath$Female))
meta.matched.Horvath$Female<-as.numeric(meta.matched.Horvath$Female)
meta.matched.Horvath$Tissue<-gsub("Cortex", "Brain FCTX", gsub("Liver", "Liver", meta.matched.Horvath$Tissue))
identical(colnames(betas.matched), meta.matched.Horvath$Sample_ID) #TRUE
write.table(meta.matched.Horvath,"Meta.matched.Horvath.csv", row.names=F, sep=",")
```

## E. Analysis of output from online epigenetic age calculator

```{r}
## load outputs (as csv files)
cortex.output<-read.csv("CortexDat_DNAm_Age_Analysis_Output.csv", header=T) # cortex dataset analysis
matched.output<-read.csv("MatchedDat_DNAm_Age_Analysis_Output.csv", header=T) # matched dataset analysis



### analyze cortex-only dataset output
## look at DNAm Age
ggplot(cortex.output, aes(Disease, DNAmAge))+
  geom_boxplot(aes(fill=Disease),color="black", outlier.shape=NA)+
  geom_point(shape=21, fill="grey",colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Disease Status")+ylab("Predicted DNA Methylation Age") +ggtitle("Epigenetic Age Analysis of HD cases vs controls cortex samples") + scale_fill_manual(values=c("#d53e4f", "#3288bd"), guide=FALSE) 
DNAmage.HD<-subset(cortex.output$DNAmAge, cortex.output$Disease=="HD")
DNAmage.Control<-subset(cortex.output$DNAmAge, cortex.output$Disease=="Control")
wilcox.test(DNAmage.HD, DNAmage.Control, paired=FALSE) # p-value = 0.6282 (Mann Whitney U test)

## look at Age Acceleration Residual
ggplot(cortex.output, aes(Disease, AgeAccelerationResidual))+
  geom_boxplot(aes(fill=Disease),color="black", outlier.shape=NA)+
  geom_point(shape=21, fill="grey",colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Disease Status")+ylab("Age Acceleration Residual") +ggtitle("Epigenetic Age Analysis of HD cases vs controls cortex samples") + scale_fill_manual(values=c("#d53e4f", "#3288bd"), guide=FALSE) 
AA.HD<-subset(cortex.output$AgeAccelerationResidual, cortex.output$Disease=="HD")
AA.Control<-subset(cortex.output$AgeAccelerationResidual, cortex.output$Disease=="Control")
wilcox.test(AA.HD, AA.Control, paired=FALSE) # p-value = 0.8357 (Mann Whitney U test)

## scatterplot of DNAmAge and Chronological Age (Reported)
ggplot(cortex.output, aes(Age, DNAmAge)) + geom_point(aes(color = AgeAccelerationResidual)) + geom_smooth(method="lm") +
  theme_bw()+xlab("Chronological Age")+ylab("DNA Methylation Age") 

# look at epigenetic age in cortex samples of HD cases only
load("HTT_Leavitt.cortex.cor.final.RData") 
HD.meta<-subset(pData(HTT_Leavitt.cortex.cor), pData(HTT_Leavitt.cortex.cor)$Disease_status=="HD")
identical(rownames(HD.meta),as.character(HD.output$Sample_ID)) # TRUE
colnames(HD.meta)
HD.output$Age_Onset<-HD.meta$Age_Onset
## scatterplot of DNAmAge and Chronological Age (Reported)
ggplot(HD.output, aes(Age_Onset, AgeAccelerationResidual)) + geom_point(aes(fill = DNAmAge), shape=21, size=5)  +
  theme_bw()+xlab("Age of Disease Onset")+ylab("Age Acceleration Residual") 

### analyze matched-only dataset output
## look at DNAm Age
ggplot(matched.output, aes(Tissue, DNAmAge))+
  geom_boxplot(aes(fill=Tissue),color="black", outlier.shape=NA)+
  geom_point(aes(shape = Disease),  fill="grey",colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Tissue Type")+ylab("Predicted DNA Methylation Age") +ggtitle("Epigenetic Age Analysis of Cortex vs Liver in Matched Dataset") + scale_fill_manual(values=c("#f1a340", "#998ec3"), guide=FALSE) 
DNAmage.Cortex<-subset(matched.output$DNAmAge, matched.output$Tissue=="Cortex")
DNAmage.Liver<-subset(matched.output$DNAmAge, matched.output$Tissue=="Liver")
wilcox.test(DNAmage.Cortex, DNAmage.Liver, paired=TRUE) # p-value = 1 (Wilcoxon Signed Rank Paired Test)

## look at Age Acceleration Residual
ggplot(matched.output, aes(Tissue, AgeAccelerationResidual))+
  geom_boxplot(aes(fill=Tissue),color="black", outlier.shape=NA)+
  geom_point(aes(shape = Disease),  fill="grey",colour="black", size=4, position=position_jitter(w=0.2))+
  theme_bw()+xlab("Tissue Type")+ylab("Age Acceleration Residual") +ggtitle("Epigenetic Age Analysis of Cortex vs Liver in Matched Dataset") + scale_fill_manual(values=c("#f1a340", "#998ec3"), guide=FALSE) 
AA.Cortex<-subset(matched.output$AgeAccelerationResidual, matched.output$Tissue=="Cortex")
AA.Liver<-subset(matched.output$AgeAccelerationResidual, matched.output$Tissue=="Liver")
wilcox.test(AA.Cortex, AA.Liver, paired=TRUE) # p-value = 1 (Wilcoxon Signed Rank Paired Test)

## scatterplot of DNAmAge and Chronological Age (Reported)
ggplot(matched.output, aes(Age, DNAmAge)) + geom_point(aes(color = AgeAccelerationResidual)) + geom_smooth(method="lm") +
  theme_bw()+xlab("Chronological Age")+ylab("DNA Methylation Age") 
```



